Molecular dynamics simulation of synchronization of a driven particle

### 2021-04-23 manuscript/manuscript_ajp_version3.tex
revisions to text, fig 4

The model discussed in this paper
resembles an externally forced pendulum.
%damping
%transition
%%DM - focus more on single particle with environment
A single pendulum 
can swing in time with an %be driven or damped by
driving force
like  %a skateboarder in a half-pipe or
a pushed child on a swing,
and may be damped
by mechanical friction.
%of gravity to return
%to an equilibrium position,
%or  to decreasing amplitude
% or air resistance.
We focus on 
synchronized phase-locking or mode-locking,
which 
first appeared in the scientific literature
in 1665 with 
Huygens'  experiments on the
motions of %synchronized 
pendula in wall-mounted clocks.\cite{Bennett2002}
A locked-mode is an integer frequency ratio
in coupled oscillators. %Clock pendula swing in phase
%when coupled through shared vibrations.
In Huygens' clocks,
the mode-locked pendula were observed to swing at the same rate
in the same direction (1:1 mode) or
opposite directions ($-1$:1 mode).
Higher modes might be achieved
by altering the pendula lengths - 
a 2:1 mode occurs when a simple pendulum
swings with twice the period
of another one quarter the length.
%DM transition sentence
%Many subsequent 
Simulations and laboratory experiments
have revealed 
mode-locked synchronization
in a diverse range of 
oscillation phenomena.

%Single oscillators 
%can demonstrate a variety of synchronized behaviors.
Colloid particles in complex environments are an excellent
model system to explore %these 
the dynamics of mode-locking. %synchronization.
In experiments %DM moved the ``in experiments'' to second sentence
typical colloids are   
plastic spheres suspended in
de-ionized water or silica beads suspended in organic solvent.
Colloids can be trapped
with 
radiation pressure from 
a laser beam.\cite{Ashkin1997} 
A colloid in such a confining region,
an optical trap, oscillates
about a equilibrium point.
A colloid centered in an optical trap is 
uniformly bombarded by photons. 
Off-center colloids 
experience a net force
due to uneven photon collisions across
the particle surface.
Depending on the 
location of the particle in the trap,
the radiation pressure either moves colloids toward the center 
or ejects it from the trap.
Diffraction gratings can create
more complex light environments, 
such as periodic patterns of minima
suitable for synchronization studies.\cite{Grier2003}

Colloids have been
shown to 
provide an ideal system to 
control and study
phase-locked motions
%are useful for
%understanding synchronization
at a single particle level.\cite{Juniper2015,Juniper2017}
% such as synchronized mode-locking can be studied with
%An advantage of colloid studies
A driven colloid
can demonstrate
synchronized
motion within one trap,
or synchronized hopping between multiple traps. 
While many systems oscillate,
trapped colloids are useful for 
providing controlled 
measurements of the
step-by-step dynamics of particle motion.
Since
colloids are large 
and move slowly, 
%colloids are considered model systems
particle positions 
can be measured in real time 
with an optical camera.\cite{Pertsinidis2001}
%Light is a
%tool for manipulating the colloidal environment,
%which can alter synchronization patterns.
%Synchronization patterns can be altered
%using light to manipulate
%the colloidal environment.
Simulations provide complementary
information,
where environmental and driving parameters may be precisely
and quickly modified to explore
the many oscillation patterns.
%the dynamics from external
In the following study,
we confine colloids  
to move along a periodic light pattern
with minima and maxima defined by a sine wave,
a washboard potential energy substrate.

@@

The confining force on the particle $i$
 is given by
 \begin{equation}
 \vec{F}_{\ell}(\vec{r}) = - \vec \nabla U(\vec{r}),
 \label{eq:dudr}
 \end{equation}
where $\ell$ denotes the landscape.
 In Fig.~\ref{fig:1_landscape}(b) we plot  
 $U(y)$ to illustrate how the magnitude
 $|\vec{F}_{\ell}|$ is calculated from the particle position $y$.

@@

The inertia of the 
small particle is reduced by interactions
with the fluid particles.\cite{Purcell1977}
We assume 
colloids are
overdamped
so the particle does not accelerate,
that is, it is suspended in a continuous viscous fluid
that dissipates energy supplied externally. 
Newton's second law for the particle
is simplified
by the assumption that $\vec{a}$ is zero. 
The overdamped equation of motion for
the velocity $\vec{v}$ of 
an isolated particle is
\begin{equation}
  \eta \vec{v} = \vec{F}_{\ell}(\vec{r}) + \vec{F}_{\rm d}(t),
    \label{eq:motion}
\end{equation}
with friction coefficient $\eta = 1$ in units of $v_0 / F_0$.
The term $-\eta \vec{v}$
is a drag force that models
energy dissipation due to the fluid. 
We discuss  models for
a sphere moving through a fluid in 
Problem~\ref{ex:reynolds}.

@@

  To study synchronization patterns, 
  it is useful to compare
  mode-locked quantities 
  in a two-dimensional phase plot. 
  For a driven pendulum confined to a single potential well,
  an appropriate
  phase space is the particle velocity $v_y$ versus the position $y$.  
  For a particle driven 
  through multiple identical wells 
  we define phase variables 
  to account for the net increase in the position.
  The phase position is
  \begin{equation}
    \phi(t) = 2\pi [y(t)-\langle v_y \rangle t]/\lambda,
  \end{equation}
  centered about the average particle displacement $\langle v_y \rangle t$
  and normalized by the substrate period $\lambda$.\cite{Juniper2015}
  The phase velocity is
  \begin{equation}
    \dot{\phi}(t) =2\pi [v_y(t)-\langle v_y \rangle] /\lambda.  
  \end{equation}
For a zero landscape force, 
  the phase velocity  is zero when $F_{\rm d}(t) = F_{\rm dc}$.
If the system
  is strictly mode-locked,
  the particle velocity
  recurs at a particular spatial location, and 
a closed loop appears in  
  phase space. 
  A 1:1 mode appears as a circle or oval. 
  Nodes appear 
  for higher modes,
  sometimes forming figure-eights
  or other recognizable patterns.
  A system that is nearly phase locked
  will appear as an almost closed loop.
  Such quasiperiodic systems are
  not fully synchronized
  so the position-velocity relation
  shifts in time as in Fig.~\ref{fig:4_phase}.

  In Fig.~\ref{fig:4_phase}
  we plot $\dot{\phi}(t)$ versus $\phi(t)$
  for increasing 
  $F_{\rm dc}$, 
  with the remaining parameters fixed as in Fig.~\ref{fig:2_Fd_vy_time}.
For
  $F_{\rm dc} = 0.04$
  the phase plot   in Fig.~\ref{fig:4_phase}(a)  is an
  asymmetric curve.
  A tail appears
  due to the initial transient
  motion of the particle.
  The particle is confined to a single
  substrate minima,
  and has no net velocity.
  The asymmetry is caused by the bias induced by $F_{\rm dc}$.
  In Fig.~\ref{fig:4_phase}(b)
  with $F_{\rm dc} = 0.07$
  the phase loop is a symmetric triangular shape,
  indicating a 1:1 match between the
  particle motion and velocity consistent with 
  Fig.~\ref{fig:2_Fd_vy_time}(a).
  As $F_{\rm dc}$ is
  increased,
  nodes form in the phase diagram, which occur
  due to repeated values
  of $\dot{\phi}$ over multiple phase positions.
  In Fig.~\ref{fig:4_phase}(c)
  with $F_{\rm dc} = 0.1$
  two nodes form.
  The particle moves a distance $2\lambda$
  during one time period.
For $F_{\rm dc} = 0.125$ as in Fig.~\ref{fig:4_phase}(d),
  three nodes form as the particle moves across $3\lambda$.

@@

Our simulations reproduce the experiments and simulations presented in 
Juniper {\it et al.} \cite{Juniper2015, Juniper2017}
of 
mode locking in
driven colloids on a
periodic optical landscape.
A single particle driven across a periodic potential landscape 
synchronizes its motion 
to environmental and external forces. 
%The model
%is easy to simulate yet relevant
%to a broad range of condensed matter systems.
Our results are relevant 
to synchronization effects
in a broad range of experimental systems
including optically confined colloids,
superconductors with periodic pinning arrays, 
and the charge and spin of atomic systems.

Colloids are 
relatively easy to 
manipulate and image in experiments,
making ideal proxies 
for systems 
such as cold atoms or electron gases.\cite{Grier2003}
%DM FIX FOLLOWING SENTENCE
Dynamical mode-locking 
is 
observed in technologies such as 
quantum electronic
devices as 
stepped regions in the relation between current and voltage,
where the voltage is the analog of the external driving force
and current is the analog of particle velocity.
%DM FIX FOLLOWING SENTENCE
Known as Shapiro steps, 
these mode-locked or phase-locked currents
%arise 
%due to applied ac voltages
were first
observed in single Josephson junctions\cite{Shapiro1963, Golubov2004}.
%coupled arrays of junctions.\cite{Benz1990}
%DM END FIX
Shapiro steps vary in width depending on the strength of the
applied ac forces,
and are observed in a variety of ac and dc driven systems
displaying
non-Ohmic behavior in voltage-current curves,
including
charge density waves, spin density waves
and superconducting vortices in landscapes 
engineered with periodic patterns of pinning sites.\cite{Reichhardt2000}
Mode-locking is a useful probe 
of complex quantum mechanical systems
because the motions of individual particles can  be inferred only
from other measurements.
%insert final sentence summarizing all the things.

@@

  \noindent To calculate the position of the particle
  as a function of time,
  we
  numerically integrate the equation of motion,
  Eq.~(\ref{eq:motion}),
  using
  the standard definition of velocity
  $\vec{v} = d\vec{r}/dt$ 
  using the 
  Euler method.
  Equation~(\ref{eq:motion}) provides
  a direct calculation for the particle velocity
  from the net force on the particle,
  as demonstrated in Problem~\ref{ex:n2l}.

@@

  \item[(a)] {\bf Initialization.}
    For convenience,
    we define a Python dictionary
    to contain the simulation constants
    that could easily be passed by reference
    to subroutines.
    We have left as an exercise  the addition
    of the remaining parameters of the system.
    \begin{verbatim}
def set_parameters():           
    ```set simulation parameters...'''
    #declare the dictionary     
    dict={}  
    dict[`dt'] = 0.1 #time step in simulation units
    #control the oscillating component of driving force
    dict[`F_AC'] = 0.07 #amplitude of force oscillation
    dict[`freq'] = 0.01 #frequency of force oscillation
    #...                #left as an exercise to the reader
    return dict
    \end{verbatim}

>>>>

The model discussed in this paper
resembles an externally forced pendulum.
%damping
%transition
%%DM - focus more on single particle with environment
A single pendulum 
can swing in time with an %be driven or damped by
driving force
like  %a skateboarder in a half-pipe or
a pushed child on a swing,
and may be damped
by mechanical friction.
%of gravity to return
%to an equilibrium position,
%or  to decreasing amplitude
% or air resistance.
We focus on 
synchronized phase-locking or mode-locking,
which 
first appeared in the scientific literature
in 1665 with 
Huygens'  experiments on the
motions of %synchronized 
pendula in wall-mounted clocks.\cite{Bennett2002}
A locked-mode is an integer frequency ratio
in coupled oscillators. %Clock pendula swing in phase
%when coupled through shared vibrations.
In Huygens' clocks,
the mode-locked pendula were observed to swing at the same rate
in the same direction (1:1 mode) or
opposite directions ($-1$:1 mode).
Higher modes might be achieved
by altering the pendula lengths - 
a 2:1 mode occurs when a simple pendulum
swings with twice the period
of another one quarter the length.
%DM transition sentence
%Many subsequent 
Simulations and laboratory experiments
have revealed 
mode-locked synchronization
in a diverse range of 
oscillation phenomena.
%Single oscillators 
%can demonstrate a variety of synchronized behaviors.
%Colloid particles in optical traps are an excellent
%model system to explore %these 
%the dynamics of mode-locking. %synchronization.

Colloids have been
shown to 
provide an ideal system to 
control and study
phase-locked motions
%are useful for
%understanding synchronization
at a single particle level.\cite{Juniper2015,Juniper2017}
% such as synchronized mode-locking can be studied with
%An advantage of colloid studies
In experiments %DM moved the ``in experiments'' to second sentence
typical colloids are   
plastic spheres suspended in
de-ionized water or silica beads suspended in organic solvent.
Colloids can be trapped
with 
radiation pressure from 
a laser beam.\cite{Ashkin1997} 
%A colloid in such a confining region,
%an optical trap,
A colloid centered in an optical trap is 
uniformly bombarded by photons
and oscillates
about a equilibrium point.
Off-center colloids 
experience a net force
due to uneven photon collisions across
the particle surface.
%Depending on the 
%location of the particle in the trap,
%the radiation pressure either moves colloids toward the center 
%or ejects it from the trap.
Diffraction gratings can be used to create
more complex light environments, 
such as periodic patterns of minima
suitable for synchronization studies.\cite{Grier2003}  
A driven colloid
can demonstrate
synchronized
motion within one trap,
or synchronized hopping between multiple traps. 
While many systems oscillate,
trapped colloids are useful for 
providing controlled 
measurements of the
step-by-step dynamics of particle motion.
Since
colloids are large 
and move slowly, 
%colloids are considered model systems
particle positions 
can be measured in real time 
with an optical camera.\cite{Pertsinidis2001}
%Light is a
%tool for manipulating the colloidal environment,
%which can alter synchronization patterns.
%Synchronization patterns can be altered
%using light to manipulate
%the colloidal environment.
Simulations provide complementary
information,
where environmental and driving parameters may be precisely
and quickly modified to explore
the many oscillation patterns.
%the dynamics from external
In the following study,
we confine colloids  
to move along a periodic pattern
with minima and maxima defined by a sine wave.
%a washboard potential energy substrate.

@@

The confining force on the particle %$i$ DM removed
 is given by
 \begin{equation}
 \vec{F}_{\ell}(\vec{r}) = - \vec \nabla U(\vec{r}),
 \label{eq:dudr}
 \end{equation}
where $\ell$ denotes the landscape.
 In Fig.~\ref{fig:1_landscape}(b) we plot  
 $U(y)$ to illustrate how the magnitude
 $|\vec{F}_{\ell}|$ is calculated from the particle position $y$.

@@

The inertia of the 
%small
particle is reduced by interactions
with the fluid particles.\cite{Purcell1977}
We assume 
colloids are
overdamped
so the particle does not accelerate,
that is, it is suspended in a continuous viscous fluid
that dissipates energy supplied externally. 
Newton's second law for the particle
is simplified
by the assumption that $\vec{a}$ is zero. 
The overdamped equation of motion for
the velocity $\vec{v}$ of 
an isolated particle is
\begin{equation}
  \eta \vec{v} = \vec{F}_{\ell}(\vec{r}) + \vec{F}_{\rm d}(t),
    \label{eq:motion}
\end{equation}
with friction coefficient $\eta = 1$ in units of $v_0 / F_0$.
The term $-\eta \vec{v}$
is a drag force that models
energy dissipation due to the fluid. 
We discuss  models for
a sphere moving through a fluid in 
Problem~\ref{ex:reynolds}.

@@

  To study synchronization patterns, 
  it is useful to compare
  mode-locked quantities 
  in a two-dimensional phase plot. 
  For a driven pendulum confined to a single potential well,
  an appropriate
  phase space is the particle velocity $v_y$ versus the position $y$.  
  For a particle driven 
  through multiple identical wells 
  we define phase variables 
  to account for the net increase in the position.
  The phase position is
  \begin{equation}
    \phi(t) = 2\pi [y(t)-\langle v_y \rangle t]/\lambda,
  \end{equation}
  centered about the average particle displacement $\langle v_y \rangle t$
  and normalized by the substrate period $\lambda$.\cite{Juniper2015}
  The phase velocity is
  \begin{equation}
    \dot{\phi}(t) =2\pi [v_y(t)-\langle v_y \rangle] /\lambda.  
  \end{equation}
  %DM removed
  %For a zero landscape force, 
  %the phase velocity  is zero when $F_{\rm d}(t) = F_{\rm dc}$.
If the system
  is strictly mode-locked,
  the particle velocity
  recurs at a particular spatial location, and 
a closed loop appears in  
  phase space. 
  A 1:1 mode appears as a circle or oval. 
  Nodes appear 
  for higher modes,
  sometimes forming figure-eights
  or other recognizable patterns.
  A system that is nearly phase locked
  will appear as an almost closed loop.
  Such quasiperiodic systems are
  not fully synchronized
  so the position-velocity relation
  shifts in time as in Fig.~\ref{fig:4_phase}.

  In Fig.~\ref{fig:4_phase}
  we plot $\dot{\phi}(t)$ versus $\phi(t)$
  for increasing 
  $F_{\rm dc}$, 
  with the remaining parameters fixed as in Fig.~\ref{fig:2_Fd_vy_time}.
  %DM add
  We choose values of $F_{\rm dc}$
  corresponding to the first four modes in
  Fig.~\ref{fig:3_sweep_vyFDC},
  as marked with black circles.
  %DM end add
  For
  $F_{\rm dc} = 0.04$
  the phase plot   in Fig.~\ref{fig:4_phase}(a)  is an
  asymmetric curve.
  A tail appears
  due to the initial transient
  motion of the particle.
  The particle is confined to a single
  substrate minima,
  and has no net velocity.
  The asymmetry is caused by the bias induced by $F_{\rm dc}$.
  In Fig.~\ref{fig:4_phase}(b)
  with $F_{\rm dc} = 0.07$
  the phase loop is a symmetric triangular shape,
  indicating a 1:1 match between the
  particle motion and velocity consistent with 
  Fig.~\ref{fig:2_Fd_vy_time}(a).
  As $F_{\rm dc}$ is
  increased,
  nodes form in the phase diagram, which occur
  due to repeated values
  of $\dot{\phi}$ over multiple phase positions.
  In Fig.~\ref{fig:4_phase}(c)
  with $F_{\rm dc} = 0.1$
  two nodes form.
  The particle moves a distance $2\lambda$
  during one time period.
For $F_{\rm dc} = 0.125$ as in Fig.~\ref{fig:4_phase}(d),
three nodes form as the particle moves across $3\lambda$.
%DM add
It is possible to create
more patterns in phase diagrams are possible
by changing the driving parameters.
At higher values of $F_{\rm dc}$
additional nodes appear.
When the values of $F_{\rm ac}$ and $f$
are tuned to produce backwards hops in the substrate,
as suggested in Problem~\ref{ex:parameters}, 
negative nodes appear in $\phi_y(t)$.
%DM end add

@@

A single particle driven across a periodic potential landscape 
can synchronize its motion 
to environmental and external forces. 
The simulations described above
reproduce the experiments and simulations presented in 
Juniper {\it et al.} \cite{Juniper2015, Juniper2017}
of dynamical 
mode locking in
driven colloids on a
periodic optical landscape.
Further simulations can be extended
to include multiple particles
and more complex environments,
leading to
synchronization effects relevant 
to a broad range of experimental systems.
%The model
%is easy to simulate yet relevant
%to a broad range of condensed matter systems.

The study of motions of particles 
sliding across periodic and aperiodic surfaces 
is relevant 
to a broad range of condensed matter systems.
%beyond optically confined colloids.
Colloids are 
relatively easy to 
manipulate and image in experiments,
making ideal proxies 
for systems
more difficult to access experimentally 
such as cold atoms or electron gases.\cite{Grier2003}
Many systems can be modeled
as particles moving across a surface
including 
the propagation of patterns
in charge and spin properties of atomic systems.
%superconductors with periodic pinning arrays, and 
%DM FIX FOLLOWING SENTENCE
Dynamical mode-locking 
is 
observed in %technologies such as 
quantum electronic
devices as 
stepped regions in the relation between current and voltage,
where the voltage is the analog of the external driving force
and current is the analog of particle velocity.
%DM FIX FOLLOWING SENTENCE
Known as Shapiro steps, 
these mode-locked or phase-locked currents
%arise 
%due to applied ac voltages
were first
observed in single Josephson junctions\cite{Shapiro1963, Golubov2004}.
%coupled arrays of junctions.\cite{Benz1990}
%DM END FIX
Shapiro steps vary in width depending on the strength of the
applied ac forces,
and are observed in a variety of ac and dc driven systems
displaying
non-Ohmic behavior in voltage-current curves.
The study of 
mode-locking in colloids provides insights
into 
%is a useful probe of
complex %quantum mechanical
systems in which 
the motions of individual particles can  be inferred only
from other measurements, 
%insert final sentence summarizing all the things.
including
charge density waves, spin density waves
and superconducting vortices in landscapes 
engineered with periodic patterns of pinning sites.\cite{Reichhardt2000}

@@

  \noindent To calculate the position of the particle
  as a function of time,
  we
  numerically integrate the equation of motion,
  Eq.~(\ref{eq:motion}),
  using
  the standard definition of velocity
  $\vec{v} = d\vec{r}/dt$ 
  with the %DM sub ``with'' for using
  Euler method.
  %DM - fix Equation/Eq. discontinuity
  Eq.~(\ref{eq:motion}) provides
  a direct calculation for the particle velocity
  from the net force on the particle,
  as demonstrated in Problem~\ref{ex:n2l}.

@@

  \item[(a)] {\bf Initialization.}
    For convenience,
    we define a Python dictionary
    to contain the simulation constants
    that could easily be passed by reference
    to subroutines.
    %DM - more clearly defined task in active / command tense
    %We have left as an exercise  the addition
    Define 
    the remaining parameters of the system using the dictionary outlined below.
    \begin{verbatim}
def set_parameters():           
    ```set simulation parameters...'''
    #declare the dictionary     
    dict={}  
    dict[`dt'] = 0.1 #time step in simulation units
    #control the oscillating component of driving force
    dict[`F_AC'] = 0.07 #amplitude of force oscillation
    dict[`freq'] = 0.01 #frequency of force oscillation
    #...                #left as an exercise to the reader
    return dict
    \end{verbatim}
